A general and efficient representation of Ancestral Recombination Graphs

### 2022-07-27 paper.tex

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with \emph{nodes}
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG.
To recover
the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
simulating an ARG in this representation is
complicated by the fact that the
first event to occur may hit the lineage carrying the ancestry
to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course,
but depending on the ordering of otherwise indistinguishable
objects is generally problematic. In practice, several
methods
explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.

% it's not reasonable to assume that one type of event happens
% at a time any more.
The eARG is also limited in the types of ancestral history that can
be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough
that the number of extant lineages remains
much smaller than $N$ at all times. These approximations are
reasonable to make for small datasets, such as
the classic \citet{kreitman1983nucleotide} \emph{Drosophila melanogaster}
ADH gene dataset (comprising 11 genomes over 2.7 kb of sequence with 43 SNPs,
see the ``\nameref{ARG_inference_methods}'' section). However,
they are highly questionable in modern datasets. % absurd really
For example, many
human datasets consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
The 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype data \citep[e.g.][]{Cesarani2022}. For example,
the US dairy cattle database alone comprises more than 6 million animals with SNP array genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
This low effective population size is caused by intense selection and
staggering use of a handful of bulls across the world. The current record
holder is the bull Toystory that has produced 2.4 million doses of semen
for artificial insemination across 13 years\footnote{\url{https://www.dairyherd.com/news/world-renowned-toystory-bull-has-died}}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Similar datasets are becoming available also in other livestock species.
Notably, \citet{RosFreixedes2022} have combined genome
and SNP array data to infer recombination of genomes for 440,610 pigs within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, which is increasingly
captured in the incredibly rich datasets that are available today.

@@

Fortunately, it is straightforward to generalise the classical Griffiths
eARG encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
The only change that we require is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}C
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(d,e)$
and the interval $(x, L)$ with $(d, f)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between between positions $x$ and $y$ was inherited by
node $d$ from $e$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
$\{(x, y]\}$ with $(d,f)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

\section*{Genome ARGs}
% We're not the first to talk about ARGs and pedigrees, and we're
% not doing anything controversial.
The deep connection between ARGs and pedigrees is
well known
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of an ARG
as a \emph{data structure} in terms of these two events is problematic,
fundamentally limiting the patterns of ancestry that can be described.
Following \cite{mathieson2020ancestry} instead of focussing on events,
we define the ARG data structure in terms of \emph{genomes} and the
patterns of transmission between them.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
The Ancestral Recombination Graph defined in terms of genomes embedded
in a pedigree. (A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal (left) and maternal (right) genomes
labelled \textsf{a} to \textsf{p}. Black lines show transmission paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Two recombination breakpoints, at positions 2 and 7 mean that genomes \textsf{a}
and \textsf{c} are independent recombinants of the paternal genomes \textsf{e}
% Gregor: there is some overlap between a and c still (the middle region inherited
% from f) so, a and c are not fully independent!
and \textsf{f}: in these cases the edge annotations have been shown, indicating the
region of genome transmitted. Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker; this shows that all
genomic regions fully coalesce in genome \textsf{n} in the oldest generation.
(B) The equivalent genome ARG, where there is a single type of node (representing
a genome) along with annotations on all edges; annotations showing partial
transmission are highlighted in bold. Recombination, such as that causing the
breakpoint at position 7, is associated with two parent genomes (\textsf{e} and
\textsf{f}) and a child genome (the recombinant, \textsf{c}).
(C) The edge-annotated genome ARG in (B) fully defines the
local relationships along the genome, as shown in the local
(``marginal'') trees.
}
\end{figure}

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or gARG.
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG.
Fig.~\ref{fig-arg-in-pedigree}A shows the genomes
embedded in an inbred diploid pedigree, in which individuals occur at
discrete generations and carry two genomes. Each individual (labelled
$I_1$ to $I_8$) therefore has a paternal genome (on the left) and a maternal one
(on the right), although the pedigree is only shown for concreteness, and
there is no inherent limitation on mating system, ploidy, or age structure.
A genome is inherited from one of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination. Fig.~\ref{fig-arg-in-pedigree}B shows the resulting gARG,
depicting the ancestry of the sample genomes \textsf{a}--\textsf{d}. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-arg-in-pedigree}C).

@@

The final point deals with the way in which recombination is represented, and is
closely linked to which genomes are chosen as the nodes in a gARG.
We could potentially choose genomes at any point in the life cycle: for instance,
we could chose to represent gametes as ARG nodes. However, the most intuitive
choice is where genomes represent diploid individuals in each generation, as in
Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
is not represented by a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place (e.g.
nodes \textsf{e} and \textsf{f}), and a
single resulting ``recombinant'' genome (e.g. \textsf{c}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would normally be present on the
recombination node in an eARG. For example, node \textsf{c} in
Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{f}
and positions 7 to 10 from \textsf{e}, indicating that the breakpoint is at position 7.

>>>>

The ordering of edges is a vital element of the eARG encoding.
This is because there is a fundamental ambiguity involved
in associating recombination breakpoints with nodes
rather than edges,
and the only way we can break this symmetry is to assign
some meaning to the order in which the parents are listed
(\cite{griffiths1997ancestral} refer to the ``left'' and ``right''
parents of a recombination event). This is usually described
in terms of the process in which we extract the local trees
along the genome from an eARG.
To recover
the tree at genome coordinate $x$,
we traverse the graph rootwards from the leaves. At a particular
node $u$, if it has one parent we are at a common ancestor
node and we follow that parent. If we are at a
recombination node, $u$ has two parents: if
$x$ is less than the breakpoint $\sigma(u)$ we follow
the edge to the first parent, and otherwise follow the edge
to the second parent.
This ordering requirement, while straightforward
to describe, has some practical drawbacks. For example,
% Surely the meaning is clear from the context here and we don't need
% qualify the meaning of ``ARG''
simulating an ARG in this representation is
complicated by the fact that the
first event to occur may hit the lineage carrying the ancestry
to the right of the breakpoint rather
than the left, and so we cannot emit edges as they are generated.
Such issues can be worked around, of course,
but depending on the ordering of otherwise indistinguishable
objects is generally problematic. In practice, several
methods
explicitly associate information with the outbound edges
of a recombination event
to resolve the problem~\citep{lyngso2005minimum,ignatieva2021kwarg}.

% it's not reasonable to assume that one type of event happens
% at a time any more.
The eARG is also limited in the types of ancestral history that can
be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$, and that the genome is short enough
that the number of extant lineages remains
much smaller than $N$ at all times.
While the coalescent model can be suprisingly robust to
basic violations of the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, which is increasingly
captured in the incredibly rich datasets that are available today.

Firstly, complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}. Thus, we can
expect datasets to span the full continuum from non-recombining
short genomes to very long genomes with high-levels of recombination
[rephrase, and some cites?],
all of which have high-quality whole genome data.

Secondly, the scale of modern datasets presents a major challenge
to the assumption that $n \ll N_e$.
Several human datasets now consist of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size can be substantially \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
The 1000 bull genomes project~\citep{hayes20191000}
comprises close to 7000 genomes, which are part of multi-generation pedigrees
with millions of animals and extensive SNP array genotype and phenotype
data \citep[e.g.][]{Cesarani2022}.
For example,
the US dairy cattle database alone comprises more than 6 million
animals with SNP array
genotypes\footnote{\url{https://queries.uscdcb.com/Genotype/counts.html}},
while the effective population size in modern dairy cattle breeds is
less than 100 and decreasing~\citep{MacLeod2013,Makanjuola2020}.
This low effective population size is caused by intense selection and
staggering use of a handful of bulls across the world. The current record
holder is the bull Toystory that has produced 2.4 million doses of semen
for artificial insemination across
13 years\footnote{\url{https://www.dairyherd.com/news/world-renowned-toystory-bull-has-died}}.
An extreme sign of these breeding practices is that there are only two ancestral
Y-chromosome linages present in today's US Holstein dairy breed~\citep{Yue2015}.
Similar datasets are becoming available also in other livestock species.
Notably, \citet{RosFreixedes2022} have combined genome
and SNP array data to infer recombination of genomes for 440,610 pigs within
7 multi-generation pedigrees~\citep{whalen2018,Johnsson2021,Ros-Freixedes2020}.
Effective population size in modern pig breeds is also less than 100 due to
intense selection and directed reproduction \citep{Hall2016,Porcnic2016}.

@@

Fortunately, it is straightforward to generalise the classical Griffiths
eARG encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
The only change that we require is to associate
the details of inheritance intervals with \emph{edges} rather than
as breakpoints associated with recombination events and
edge ordering rules. This is illustrated in Fig~\ref{fig-arg-data-structure}C
where this alternative edge annotation formulation is
shown. By associating the interval $(0, x)$ with edge $(d,e)$
and the interval $(x, L)$ with $(d, f)$ we make the encoding
unambiguous: we can list the edges in any order, and it
will describe the same set of local genealogical trees.
This approach is also more flexible, because any pattern
of inheritance can be described by these intervals. For
example, if a gene conversion event occurred where a segment
between positions $x$ and $y$ was inherited by
node $d$ from $e$, we would state this by associating
the intervals $\{(0, x], (y, L]\}$ with $(d,e)$ and
$\{(x, y]\}$ with $(d,f)$.
This change is straightforward, involves no loss of information
and (as we see in later sections)
greatly increases both the expressivity and computational
power of ancestral recombination graph data structures.

\section*{Genome ARGs}
% We're not the first to talk about ARGs and pedigrees, and we're
% not doing anything controversial.
The deep connection between ARGs and pedigrees is
well understood
\citep[e.g.][]{wakeley2012genetics,gusfield2014recombinatorics,
speed2015naturereviewsgenetics}
and recent discussions have illustrated the
relationship~\citep{mathieson2020ancestry,brandt2021evaluation}, % more??
showing how the coalescence and recombination events in an ARG
correspond to particular genomes in the pedigree.
As discussed in the previous section, the definition of an ARG
as a data structure in terms of these two \emph{events} is problematic,
fundamentally limiting the patterns of ancestry that can be described.
Following \cite{mathieson2020ancestry}, instead of focussing on events,
we define the ARG data structure in terms of \emph{genomes} and the
patterns of transmission between them.
% Bit of duplication here, but good to make the point about being
% motivated by real existing datasets not some academic desire to
% model complexity that isn't there.
This, coupled with
the explicit association of genome coordinates with edges, allows
us to represent even the bewilderingly complex patterns of relatedness
present in modern datasets.

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{illustrations/arg-in-pedigree}
\end{center}
\caption{\label{fig-arg-in-pedigree}
% TODO caption needs another pass. One the full section has been written
% it'll be clearer what we need to say in the caption vs the text.
An Ancestral Recombination Graph defined in terms of genomes embedded
in a pedigree. (A) Diploid individuals, visualised in a highly inbred pedigree and
labelled $I_1$ to $I_8$, contain both paternal (left) and maternal (right) genomes
labelled \textsf{a} to \textsf{p}. Black lines show transmission paths connecting
genomes in the current generation (\textsf{a} to \textsf{d}) with their ancestors.
Two recombination breakpoints, at positions 2 and 7 mean that genomes \textsf{a}
and \textsf{c} are independent recombinants of the paternal genomes \textsf{e}
% Gregor: there is some overlap between a and c still (the middle region inherited
% from f) so, a and c are not fully independent!
and \textsf{f}: in these cases the edge annotations have been shown, indicating the
region of genome transmitted. Genomes are shaded such that where, backwards in time,
they merge into a common ancestor, the merged region is darker; this shows that all
genomic regions fully coalesce in genome \textsf{n} in the oldest generation.
(B) The equivalent genome ARG, where there is a single type of node (representing
a genome) along with annotations on all edges; annotations showing partial
transmission are highlighted in bold. Recombination, such as that causing the
breakpoint at position 7, is associated with two parent genomes (\textsf{e} and
\textsf{f}) and a child genome (the recombinant, \textsf{c}).
(C) The edge-annotated genome ARG in (B) fully defines the
local relationships along the genome, as shown in the local
(``marginal'') trees.
}
\end{figure}

To distinguish from the classical event-based encodings,
we call this encoding of an ARG data
structure a ``Genome ARG'', or gARG.
Fig~\ref{fig-arg-in-pedigree} illustrates the basic elements of a gARG.
Fig.~\ref{fig-arg-in-pedigree}A shows the genomes
embedded in an inbred diploid pedigree, in which individuals occur at
discrete generations and carry two genomes (maternal and paternal).
The pedigree is shown here for concreteness, and
there is no inherent limitation on mating system, ploidy, or age structure
in the gARG encoding.
A genome is inherited from one of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination. Fig.~\ref{fig-arg-in-pedigree}B shows the resulting gARG,
depicting the ancestry of the sample genomes \textsf{a}--\textsf{d}. The graph
topology along with the genome coordinate annotations on the
edges then uniquely defines the trees at every position
along the genome (Fig.~\ref{fig-arg-in-pedigree}C).

@@

The final point deals with the way in which recombination is represented, and is
closely linked to which genomes are chosen as the nodes in a gARG.
We could potentially choose genomes at any point in the life cycle: for instance,
we could chose to represent gametes as ARG nodes. However, the most intuitive
choice is where genomes represent diploid individuals in each generation, as in
Fig~\ref{fig-arg-in-pedigree}A. In this case recombination
is not represented by a single ``recombination node'' (as in an eARG), but with
two parent genomes between which recombination takes place
(e.g.\ nodes \textsf{e} and \textsf{f}), and a
single resulting ``recombinant'' genome (e.g.\ \textsf{c}). The
edges that link this recombinant with its two parents carry the information
concerning breakpoint location that would be associated with a
recombination node in an eARG. For example, node \textsf{c} in
Fig~\ref{fig-arg-in-pedigree} inherits genome positions 0 to 7 from \textsf{f}
and positions 7 to 10 from \textsf{e}, indicating that the breakpoint is at position 7.
